A Formalization of Typed Lua

### 2015-08-18 doc/papers/dls15/paper.tex
preparing dls paper for camera-ready version

%\usepackage[compact]{titlesec}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{url}
\usepackage{color}
\usepackage{multirow}

\setlength{\textfloatsep}{10pt}

@@

In this work, we give a formal description of Typed Lua, an
optional type system for Lua, with a focus on two of its novel type system features: incremental evolution of imperative record and object types that is both lightweight and type-safe, and
the combination of flow typing, functions that return multiple values, and multiple assignment. While our type system is tailored to the features and idioms of Lua, its features can be adapted to other imperative scripting languages.
\end{abstract}

@@

Ideally, programming languages should offer programmers the
option to choose between static and dynamic typing.
\emph{Optional type systems}~\cite{bracha2004pluggable} and
\emph{gradual typing}~\cite{siek2006gradual} are two 
approaches for blending static and dynamic typing in the same
language. Their aim is to offer programmers the option
of adding type annotations where static typing is needed,
allowing the incremental migration from dynamic to static
typing. The difference between these two approaches is the
way they treat the runtime semantics of the language:
optional type systems preserve the original semantics
of a dynamically typed language, at the cost of runtime
safety, while gradual typing changes the semantics to
include more extensive runtime checking. These extra checks
ensure that
dynamically typed portions of the program do not violate
the type safety of statically typed portions.

@@

The {\tt new\_point} function builds the object piece-by-piece,
from an initially empty table. Typed Lua
lets record and object types evolve incrementally 
via assignment while not sacrificing width subtyping.

@@

In the {\bf then} branch it is guaranteed that both {\tt q}
and {\tt r} are integers, and in the {\bf else} branch it is
guaranteed that {\tt q} is {\tt nil} and {\tt r} is a string.
In Typed Lua, the return type of {\tt idiv} can be a union of
tuples describing the possible return values. Typed Lua deconstructs unions of tuples in a way that the dependencies between the types of each value in the tuple can be tracked.

@@

Table types represent some of the forms that Lua
tables can take. What a particular table type denotes
depends on its {\em tag}: the {\em unique}, {\em open}
and {\em fixed} tags all denote tables where it is
guaranteed that every key belongs to one of the {\em key
types} $F_i$. 
%We restrict key types to literal types,
%base types, and the top type to make the system simpler
%for the user, as our survey of Lua programs showed that
%this restriction reflects the way Lua tables are commonly used.

Fields are assignable by default, but can be marked
as non-assignable with the {\bf const} qualifier. Having
a field be {\bf const} is not a guarantee that its
value cannot change, as the table may have aliases with
a non-{\bf const} type for that field. Non-assignable
fields enable {\em depth subtyping} between table types.

@@

We present the typing rules as a deduction system for two typing relations, one for typing statements and another for typing expressions. Typing judgments for statements have the form $\env_{1}, \penv \vdash s, \env_{2}$, meaning that
given a type environment $\env_{1}$ and a {\em projection environment} $\penv$ we can prove that a statement $s$ produces a new type environment $\env_{2}$. Typing judgments for
expressions have the form $\env_{1}, \penv \vdash e : T, \env_{2}$, meaning that given a type environment $\env_{1}$
and a projection environment $\penv$ we can prove that an expression $e$ has type $T$ and produces a new type environment $\env_{2}$.
Only the most relevant rules for the features that we are focusing on are shown.

@@

Ravi~\cite{ravi} is an experimental Lua dialect
with an LLVM-based JIT compiler. It has a basic
optional type system to improve performance
of numerical code. This type system only supports
numbers (integers and floating point) and arrays.

@@

Several dynamically typed languages now have optional
type systems: Clojure~\cite{bonnaire-sergeant2015typed-clojure},
JavaScript~\cite{typescript}, Python~\cite{mypy}, 
and PHP~\cite{hack}. While Lua has some similarities to
all of these languages, none of these optional type
systems have the features described in this paper.
Most of these languages do not have idioms
that inspired these features; JavaScript has the
idiom of adding fields to an initially empty
record through assignment, but TypeScript sidesteps
the issue by having allowing an empty record to
have any record or object type, as the type of
missing fields is the bottom type of its type system.

Dart~\cite{dart} and Grace~\cite{black2013sg} are
two languages that have been designed from scratch
to have an optional type system, instead of being
existing dynamically typed languages with a retrofitted 
type system.

@@

We performed several
case studies to evaluate how successful we were in our goal of
providing an usable type system.
We evaluated 29 modules from 8 different case studies,
and we could give precise static types to 83\% of the 449
values that these modules export.
For half of the modules, we could give precise static types to
at least 89\% of the exports from each module.

@@

>>>>

\usepackage[compact]{titlesec}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{url}
\usepackage{color}
\usepackage{multirow}

\setlength{\textfloatsep}{8pt}

@@

In this work, we give a formal description of Typed Lua, an
optional type system for Lua, with a focus on two of its novel type system features: incremental evolution of imperative record and object types that is both lightweight and type-safe, and {\em projection types}, a combination of flow typing, functions that return multiple values, and multiple assignment. While our type system is tailored to the features and idioms of Lua, its features can be adapted to other imperative scripting languages.
\end{abstract}

@@

Ideally, programming languages should offer programmers the
option to choose between static and dynamic typing.
\emph{Optional type systems}~\cite{bracha2004pluggable} and
\emph{gradual typing}~\cite{siek2006gradual} are two 
approaches for blending static and dynamic typing in the same
language. Their aim is to offer programmers the option
of adding type annotations where static typing is needed,
allowing the incremental migration from dynamic to static
typing.

The difference between these two approaches is the
way they treat the runtime semantics of the language:
optional type systems have the {\em trace preservation}
property~\cite{strongscript}, where type annotations
do not influence runtime behavior, and the original
semantics of the dynamically typed language is preserved,
 while gradual typing changes the semantics to
include more extensive runtime checking in the
interface between typed and untyped code. 
In an optionally-typed language, dynamically typed code
can lead to type errors in statically typed code, even
if the optional type system is sound. In a gradually-typed
language, the extra runtime checks ensure that 
this does not occur.
 The decision
between the two approaches is not binary, and languages
can have both optional and gradual types~\cite{liketypes,strongscript}.

@@

The {\tt new\_point} function builds the object piece-by-piece,
from an initially empty table. Typed Lua
lets record and object types evolve incrementally 
via assignment while not sacrificing width subtyping,
where we can form a supertype of a table type by ``forgetting'' fields.

@@

In the {\bf then} branch it is guaranteed that both {\tt q}
and {\tt r} are integers, and in the {\bf else} branch it is
guaranteed that {\tt q} is {\tt nil} and {\tt r} is a string.
In Typed Lua, the return type of {\tt idiv} can be a union of
tuples describing the possible return values. Typed Lua deconstructs unions of tuples into {\em projection types}, and these types track
the dependencies between the types of each value in the tuple.

@@

Table types represent some of the forms that Lua
tables can take. What a particular table type denotes
depends on its {\em tag}: the {\em unique}, {\em open}
and {\em fixed} tags all denote tables where it is
guaranteed that every key belongs to one of the {\em key
types} $F_i$.

%We restrict key types to literal types,
%base types, and the top type to make the system simpler
%for the user, as our survey of Lua programs showed that
%this restriction reflects the way Lua tables are commonly used.

Fields are assignable by default, but can be marked
as non-assignable with the {\bf const} qualifier in its key type.
Having a field be {\bf const} is not a guarantee that its
value cannot change, as the table may have aliases with
a non-{\bf const} type for that field. Non-assignable
fields enable {\em depth subtyping} between table types: these
fields are covariant on the types of their values, while
assignable fields are always invariant.

@@

We present the typing rules as a deduction system for two typing relations, one for typing statements and another for typing expressions. Typing judgments for statements have the form $\env_{1}, \penv \vdash s, \env_{2}$, meaning that
given a type environment $\env_{1}$ and a {\em projection environment} $\penv$ we can prove that a statement $s$ produces a new type environment $\env_{2}$. Typing judgments for
expressions have the form $\env_{1}, \penv \vdash e : T, \env_{2}$, meaning that given a type environment $\env_{1}$
and a projection environment $\penv$ we can prove that an expression $e$ has type $T$ and produces a new type environment $\env_{2}$.
Only the most relevant rules are shown. The first author's Ph. D. thesis has the full set of rules~\cite{maidl2015tl}.

@@

@@

Several dynamically typed languages now have optional
type systems: Clojure~\cite{bonnaire-sergeant2015typed-clojure},
JavaScript, Python~\cite{mypy}, 
and PHP. While Lua has some similarities to
all of these languages, none of these optional type
systems have the features described in this paper.
Most of these languages do not have idioms
that inspired these features; JavaScript has the
idiom of adding fields to an initially empty
record through assignment, but TypeScript sidesteps
the issue by having allowing an empty record to
have any record or object type, as the type of
missing fields is the bottom type of its type system.

Grace~\cite{black2013sg} and Dart are
two languages that have been designed from scratch
to have an optional type system, instead of being
existing dynamically typed languages with a retrofitted 
type system.

@@

We performed several
case studies to evaluate how successful we were in our goal of
providing an usable type system.
We evaluated 29 modules from 8 different case studies,
and we could give precise static types to 83\% of the 449
values that these modules export.
For half of the modules, we could give precise static types to
at least 89\% of the exports from each module. The methodology,
detailed results and analysis of this evaluation is part
of the first author's Ph. D. thesis~\cite{maidl2015tl}.

@@

\acks

The authors thank the anonymous reviewers for their feedback.
Work on Typed Lua has been supported by grants from Google
Summer of Code and CNPq, the Brazilian Research Council.
